Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-05-18 paper/letter.tex
first pass

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
features in its Lyman-$\alpha$ emission thad had evaded theoretical
interpretation so far. 
We show that these special features, a symmetric triple peaked
emission line, can be explained by gas bulk rotation.  
We find that the Lyman-$\alpha$ emission region in \tol\ should
have a rotational velocity of $V_{r}=300$ km s$^{-1}$ and a neutral
Hydrogen column density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Using archival observational information about \tol\ we find that the
diameter for that emission region diameter should be in the range of
$110$ pc $<D<$ $340$ pc and its total dynamical mass   should be
between $2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times
10^{9}$M$_{\odot}$.  
This dynamical mass is at least $16\pm 9$ times
larger than the neutral mass hydrogen.
We argue that a possibility to explain the excess in
dynamical mass is the presence of a super-massive black hole. }

@@

Radio surveys of the 21cm line have put an upper limit to the neutral
hydrogen mass in \tol of $M<2.65\times 10^{8}$ M$_{\odot}$. 
Together with our velocity and column density results, under the
assumption of spherical symmetry, we can constrain the total dynamical
mass of the HI region to be in the range $2.1\times 10^{9}$M$_{\odot}$
$<M_D<$  $6.6\times 10^{9}$M$_{\odot}$ and the diameter of HI region
should have a diameter of $0.11 < D/\mathrm{kpc}<0.34$. 
This makes the
dynamical mass  at least $7$ to $25$ times larger than the H$I$ mass.

We consider three possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass can be constrained by measurements of the 
stellar continuum, this can only represent $xx$M$_{\odot}$ of the total
contribution. 
Considering a standard dark matter halo, it can only constributio to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open  possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

A future observational test of this possibility in \tol\ would require
integral field unit measurements spatially resolving its spatial
extent. 
\tol\ spans a region of $\approx 4$ arcminutes. 
An instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a detailed mapping of different
ionization lines to infer a kinematic map.

Why is triple peaked emission not common in  Lyman-alpha emitting
gaalaxies?   
The ubiuquity of galaxy outflows overimposed to the rotation
transforms the line into the common double or single peaked
line.
Therefore, the absence of HI outflows in \tol\ is another special
feature that permitted a symmetric triple peaked lya line to emerge.

The implications of our finding for the study of LAEs, including the
first generation of galaxies are manifold. 
First of all, it demonstrated the importance of including rotation as
theoretical feature to interpret observed spectra. 
Secondly, it provides a motivation to explore under what conditions
rotation dominated dynamics are expected in comparison with outflow
dominated dynamics. 
Finally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistend with future observational
kinematic maps, it would provide a new information to test and probe
theories on the coevolution of galaxies and black holes in the first
generation of galaxies.

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MG$^3$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
features in its Lyman-$\alpha$ emission thad had evaded theoretical
interpretation so far. 
We show that these special features, a symmetric triple peaked
emission line, can be explained by gas bulk rotation.  
We find that the Lyman-$\alpha$ emission region in \tol\ should
have a rotational velocity of $V_{r}=300$ km s$^{-1}$ and a neutral
Hydrogen column density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Using archival observational information about \tol\ we find that the
diameter for that emission region diameter should be in the range of
$110$ pc $<D<$ $340$ pc and its total dynamical mass   should be
between $2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times
10^{9}$M$_{\odot}$.  
This dynamical mass is at least $16\pm 9$ times
larger than the neutral mass hydrogen.
We argue that a possibility to explain the excess in
dynamical mass is the presence of a super-massive black hole. }

@@

Radio surveys of the 21cm line have put an upper limit to the neutral
hydrogen mass in \tol\ of $M<2.65\times 10^{8}$ M$_{\odot}$. 
Together with our velocity and column density results we constrain the
total dynamical mass of the HI region to be in the range $2.1\times
10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times 10^{9}$M$_{\odot}$ and the
diameter of HI region should have a diameter of $0.11 <
D/\mathrm{kpc}<0.34$.  
This makes the dynamical mass  at least $7$ to $25$ times larger than
the neutral Hydrogen mass.

We consider three possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass can be constrained by measurements of the 
stellar continuum, this can only represent $xx$M$_{\odot}$ of the total
contribution. 
Considering a standard dark matter halo, it can only constribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

A future observational test of this possibility in \tol\ would require
integral field unit measurements spatially resolving its spatial
extent. 
\tol\ spans a region of $4$ arcminutes. 
An instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a detailed mapping of different
ionization lines to infer a kinematic map.

Why is triple peaked emission not common in Lyman-alpha emitting
gaalaxies?   
The ubiquity of galaxy outflows overimposed to the rotation
transforms the line into the common double or single peaked
line.
Therefore, the absence of HI outflows in \tol\ is another special
feature that permitted a symmetric triple peaked lya line to emerge.

The implications of our finding for the study of LAEs, including the
first generation of galaxies are manifold. 
First of all, it demonstrates the importance of including rotation as
theoretical feature to interpret observed spectra. 
Secondly, it provides a motivation to explore under what conditions
rotation dominated dynamics dominate over outflow
dominated dynamics. 
Finally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistend with future observational
kinematic maps, it would provide a new information to test and probe
theories on the coevolution of galaxies and black holes in the first
generation of galaxies.
